---
title: "Homework 6"
author: "Jessica Schaub"
date: "October 31, 2018"
output: 
  github_document:
    toc: yes
always_allow_html: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This assignment has the purpose of pulling together all concepts of data wrangling that we have learned until now.
We must select and perform two tasks from a list.
For more information, please see the [assignment](http://stat545.com/Classroom/assignments/hw06/hw06.html).

# Assignment

I have selected the following two tasks:

1. __Character Data__: Read the [Strings](https://r4ds.had.co.nz/strings.html) chapter of R for Data Science and complete the exercises.
2. __Work with the Candy Data__: Explore the [2015 candy data](https://boingboing.net/2015/10/31/the-candy-hierarchy-2015-your.html) from Cohen & Ng.

## Getting Ready

Loading the requires packages for the tasks.

```{r}
#tidyverse
suppressPackageStartupMessages(library(tidyverse))
#knitr
suppressPackageStartupMessages(library(knitr))
#stringr
suppressPackageStartupMessages(library(stringr))
#cowplot
suppressPackageStartupMessages(library(cowplot))
```


## Task 1: Character Data

### Introduction

This task requires reading through the 'Strings' chapter from R for Data Science and completing the exercises at the end of each section.
The chapter and exercises can be found [here](https://r4ds.had.co.nz/strings.html).


The number of questions that we need to complete for this assignment was not specified, so I will select a subset of questions that will allow me to practice the various applications of regex.

### String basics

**1. In code that doesn’t use stringr, you’ll often see `paste()` and `paste0()`. What’s the difference between the two functions? What stringr function are they equivalent to? How do the functions differ in their handling of `NA`?**

The difference is the default for `sep`.
`paste()` has a default value `sep = " "` if nothing is specified. 
`paste0()` has a default value `sep = ""` if nothing is specified.

```{r}
paste("Je", "ssi", "ca")
paste0("Je", "ssi", "ca")
```

`paste()` and `paste0()` don't handle NA's well. 

```{r}
a <- c("a", NA, NA)
b <- c("x", "y", NA)

paste(a, b)
paste0(a, b)
```

This default output of "NAy" is not a result I can imagine being useful in many situations. However, a more intuitive way to deal with NA's in paste does not really exist yet.

**2. In your own words, describe the difference between the `sep` and `collapse` arguments to `str_c()`.**

`str_c()` can concatenate multiple vectors together.
`sep` controls how the elements of the different vectors are separated once they are joined.
`collapse` controls how the final concatenated vectors will be collapsed down into one final string.
Without `collapse`, the output will be the same length as the vectors that were input.
With `collapse`, the length will be 1.

```{r}
#Build vectors for example
vec1 <- c("x", "y", "z")
vec2 <- c("a", "b", "c")

#Show with sep
#output is the same length as the longest vector
str_c(vec1, vec2, sep = "&")

#Show with sep & collapse
#output is one string
str_c(vec1, vec2, sep = "&", collapse = "-")

```


### Matching patterns with regular expressions

**1. What patterns will the regular expression `\..\..\..` match? How would you represent it as a string?**

`\` has a special meaning, so to represent them as a regular expression, we need to escape them with `\`. 
The string would look like: `"\\..\\..\\.."`


```{r}
writeLines("\\..\\..\\..")
```

**2. Given the corpus of common words in `stringr::words`, create regular expressions that find all words that:**

  **- Start with “y”.**
  
  **- End with “x”**
  
  **- Are exactly three letters long**

```{r}
#view words
head(words, 10) %>% 
  kable()

#find words that begin with y
words %>% 
  str_view("^y", match = TRUE)

#find words that end with x
words %>% 
  str_view("x$", match = TRUE)

#find words that are only 3 letters long
words %>% 
  str_view("^...$", match = TRUE)
```

**3. Create regular expressions to find all words that:**

  **- Start with a vowel.**
  
  **- That only contain consonants.**
  
  **- End with ed, but not with eed.**
  
  **- End with ing or ise.**
  
```{r}
#find words that start with a vowel
words %>% 
  str_view("^[aeiou]", match = TRUE)

#find words that only contain consonants
words %>% 
  str_view("^[^aeiou]+$", match = TRUE)

#find words that end with "ed", but not "eed"
words %>% 
  str_view("[^e]ed$", match = TRUE)

#find words that end with "ing" or "ise"
words %>% 
  str_view("ing$|ise$", match = TRUE)
```

**4. Create regular expressions to find all words that:**

  **- Start with three consonants.**
  
  **- Have three or more vowels in a row.**
  
  **- Have two or more vowel-consonant pairs in a row.**
  
```{r}
#find words that start with 3 consonants
words %>% 
  str_view("^[^aeiou]{3}", match = TRUE)

#find words that have 3 or more vowels in a row
words %>% 
  str_view("[aeiou]{3,}", match = TRUE)

#find words that have 2 or more vowel/consonant pairs in a row
##The output is very long so I have masked it

#words %>% 
#  str_view("(([aeiou][^aeiou])|([^aeiou][aeiou])){2,}#", match = TRUE)

```

**5. Construct regular expressions to match words that:**

  **- Start and end with the same character.**
  
  **- Contain a repeated pair of letters (e.g. “church” contains “ch” repeated twice.)**
  
  **- Contain one letter repeated in at least three places (e.g. “eleven” contains three “e”s.)**
  
```{r}
#find words that start and end with the same character
words %>% 
  str_view("^(.).+\\1$", match = TRUE)

##find words that contain a repeated pair of letters
words %>% 
  str_view("(.)\\1", match = TRUE)

##find words that contain one letter repeated in at least three places
words %>% 
  str_view("(.).+\\1.+\\1.+", match = TRUE)
```


### Tools

**1. What word has the highest number of vowels? What word has the highest proportion of vowels? **

```{r}
#turn words into df
word_df <- data.frame(words)

#add columns: number of vowels, proportion of vowels
word_df <- word_df %>% 
  mutate(vowel_num = str_count(words, "[aeiou]"),
         vowel_prop = str_count(words, "[aeiou]")/str_length(words)) 

#Check our work
word_df %>% 
  head() %>% 
  kable(col.names = c("Words", "Number of Vowels", "Porportion of Vowels"))
  
#find max number of vowels
word_df %>% 
  filter(vowel_num == max(vowel_num)) %>% 
  kable(col.names = c("Words", "Number of Vowels", "Porportion of Vowels"))

#find max proportion of vowels
word_df %>% 
  filter(vowel_prop == max(vowel_prop)) %>% 
  kable(col.names = c("Words", "Number of Vowels", "Porportion of Vowels"))

```


## Task 2: Candy Data

### Introduction

This task requires loading the candy 2015 data from Cohen & Ng.
An explanation of the data can be found [here](https://boingboing.net/2015/10/31/the-candy-hierarchy-2015-your.html).
Once the data is loaded, it must be checked and cleaned for analysis. 
This is the task I have chosen to perform, so I aim to have a data set that is ready for analysis by the end.

### Load, check, and clean the data

The data were downloaded online and named "candy_data_2015" in .csv format.

```{r}
#Loading candy 2015 data from csv file
candy_all <- read_csv('candy_data_2015.csv')
```

The data has successfully been loaded to the variable `candy_all`. Let's check the data for anomalies that need to be addressed.

```{r}
# Check dimensions, variable names, and view the results
dim(candy_all)
names(candy_all)

head(candy_all, 15) %>% 
  kable()
```

Found issues:

- 124 columns seems like a lot of variables, the data is in tidy format!
- Variable names are long strings, we can simplify the naming for some variables (age, trick_or_treating)
- ID's are missing for each respondent
- Strange values for age

Let's tackle these issues one at a time.
First, we can start by giving each participant an ID number

```{r}
# insert an id for the participants
# will be inserted at the end but will fix later 
candy_all <- candy_all %>% 
  mutate(id = row_number())
```

Now let's fix some of the variable names.

```{r}
# How old are you -> age
# Timestamp -> time
# Are you going trick or treating -> trick_or_treating
candy_all <- candy_all %>% 
  select(id, 
         age = matches("old"),
         time = matches('Time'), 
         trick_or_treating = matches('treating'), 
         everything())

#Check our work
candy_all %>% 
  head(10) %>% 
  kable()
```

Then we can fix some of the strange age values.

```{r}
#remove ages > 100 and < 1
candy_all <- candy_all %>% 
  filter(str_length(age) <= 2) %>% #ages values have either 1 or 2 digits only
  mutate(age = as.integer(age)) %>% #make all ages integers
  filter(!is.na(age)) #remove NA's

#Check our work
candy_all %>% 
  head(10) %>% 
  kable()
```

Finally, we can make the data untidy for our analysis. There are two types of data represented in this data set:

  - Candy
  - Questions about other things
  
So let's split the data frame into two data frames: `candy` & `questions`.
Metadata for ID, age, timestamp, and trick_or_treating will be preserved between both data frames.

Let's build the `candy` data frame.
Since all candy values have square brackets around them, we can search for these. 
However, square brackets are found in some questions and these questions all start with 'Please'. 
We will use this to our advantage.

```{r}
# All columns with [, with the exception of questions that contain [ (these begin with 'Please')
candy <- candy_all %>% 
  select(id,
         age,
         time,
         trick_or_treating,
         contains('['),           #candy names, all contain []
         -contains('Please'))     #without (-) questions that contain []

#Check our work
candy %>% 
  names() %>% 
  kable()

```

Now we have a data frame that only has metadata and candy information.
Before we reshape the data, we can remove the unnecessary square brackets from the candy names and replace them in the `candy` data frame.

```{r}
#isolate candy names, then remove brackets
# saved to vector 'names'
names <- candy_all %>%
  select(contains('['),           
         -contains('Please')) %>% 
  names() %>% 
  str_replace(pattern = "\\[", replacement = "") %>% 
  str_replace(pattern = "\\]", replacement = "")

#use vector 'names' to replace existing variable names in candy data frame
colnames(candy)[5:length(candy)] <- names

#check our work
candy %>% 
  head(10) %>% 
  kable()
```



Let's turn the data into untidy form.

```{r}
# Reshaping to untidy form
# gather candy values by name under 'candy'
# result is how they feel, 'feeling'
candy <- candy %>% 
  gather(key = candy, value = feeling, 5:ncol(candy))

#Check our work
candy %>% 
  head(10) %>% 
  kable()
```

We have successfully reshaped our candy data!

Now let's build the `questions` data frame.

```{r}
# Create a vector of all candy names
candy_names_original <- candy_all %>% 
  select(contains('['),              #candy names, all contain []
         -contains('Please')) %>%    #remove questions that contain []
  names()    #extract just the names

#Remove all candy names
questions <- candy_all %>% 
  select(-candy_names_original)

#Check our work
questions %>% 
  names() %>% 
  kable()

```

Now we can turn the `questions` data into untidy data.

```{r}
# out question statements in variable 'question' and answer in variable 'answer'
questions <- questions %>% 
  gather(key = question, value = answer, 5:ncol(questions))

#Check our work
questions %>% 
  head(10) %>% 
  kable()
```

Let's do a final check to see if the data is in the correct structure

```{r}
str(candy)
str(questions)
```


Success!
At this point, we have dealt with the candy data in the following ways:

  - Added ID's for each participant
  - Changed the names of the metadata columns for ease of use
  - Removed strange values for age
  - Split the all of the data into two data frames:
    - `candy`, containing all candy data and metadata
    - `questions`, containing all question data and metadata
  - Removed unnecessary square brackets from candy data
  - Reshaped the tidy data to untidy data
  
It would seem the data is ready to be explored!
